SIRV-RL: Reinforcement Learning for Optimized Policy Control during Epidemiological Outbreaks in Emerging Market and Developing Economies

### 2024-01-11 paper_draft/main.tex
paper edits

In this paper we want to optimize the government policies regarding stringency. Therefore, alongside epidemiological data, we use the measures of globally comparable government responses~\cite{Hale2021}. We use the simple SIR model without vital dynamics~\cite{Hethcote1989, Hethcote2008, ALLEN2017128} as it is assumed that the time scale is short enough so that can be neglected~\cite{Cooper2020}. By lesioning the model, as opposed to proposing a new mathematical model with more specialized compartments to more accurately represent the actual environment~\cite{Bjrnstad2020, Mwalili2020}, we can effectively address the real-world conditions and propose a solution that is both effective and extendable. However, the study has limitations; the deterministic SIR model fails to account for chance in disease spread and lacks confidence intervals on results and while stochastic models incorporate chance, they are typically more challenging to analyze than their deterministic counterparts~\cite{Hethcote2008}.

In order to capture competing costs within the environment and achieve a balance between health and economic outcomes, we intend to employ reinforcement learning~\cite{Nguyen2022, Bastani2021, Song2022, Ohi2020, PADMANABHAN2021102676}. Not only does the formulation of the model deal better with competing costs, but it also offers a more transparency behind the reasoning of the decisions being made in such circumstances. When we conceptualize our problem as a reinforcement learning task, an agent is tasked with making decisions in an environment with the aim of optimizing cumulative rewards. This approach relies on trial and error, in contrast to dynamic programming, which assumes complete knowledge of the environment~\cite{Lavet2018}. We make the use of deep reinforcement learning, which is an advancement to reinforcement learning as it allows to reduce the input dimensionality and normalize the input~\cite{Lavet2018, Arulkumaran2017, Henderson_Islam_Bachman_Pineau_Precup_Meger_2018, Mnih2015}. While our model is more transparent in terms of its dependency on the reward policy defined, it has its limitations. A universal optimal policy may not suit diverse socioeconomic contexts due to variations in healthcare resources and economic vulnerabilities across countries, regions, or cities and a comprehensive consideration of decision factors, extending beyond pure reinforcement learning results is needed~\cite{Song2022, Dunn2017, Demir2006}. Also, the current model doesn't account for the herd immunity reached through vaccination which it can be extended to account for~\cite{Marinov2022}.

\subsection*{Mathematical Formulation}
In SIR-RL, we use a compartmental model to model infectious disease. Here we divide people of the population, based on whether they are yet to come into contact with an infected person (Susceptible), are infectious themselves (Infectious), or have recovered from the infection (Recovered). These compartments create the SIR model which can be represented as follows:

@@

To estimate the parameters $\beta$ and $\gamma$ for India, based on the data from January, 2020, to October, 2022, we simply define the cost to calibrate the model using the huber loss~\cite{huberloss}:

@@

\begin{equation}
  \beta_{optimal} = 0.07565820733562756
  \label{eq:beta_optimal_without_lockdown}
\end{equation}
\begin{equation}
  \gamma_{optimal} = 0.046594691406826524
  \label{eq:gamma_optimal_without_lockdown}
\end{equation}
\begin{equation}
  R_{0_{optimal}} = \frac{\beta_{optimal}}{\gamma_{optimal}} = 1.6237516560640421
  \label{eq:r0_without_lockdown}
\end{equation}
\begin{equation}
  cost = L_{\delta = 1}(S, \hat{S}) + L_{\delta = 1}(I, \hat{I}) + L_{\delta = 1}(R, \hat{R}) = 69949738.9977647
  \label{eq:cost_without_lockdown}
\end{equation}
See figure \ref{fig:sir_without_lockdown_IND} to see how the model compares with the real data.

@@

\begin{equation}
  \frac{dS}{dt} = -\beta(t) \frac{S I}{N}
  \label{eq:S_with_lockdown}
\end{equation}
\begin{equation}
  \frac{dI}{dt} = \beta(t) \frac{S I}{N} - \gamma I
  \label{eq:I_with_lockdown}
\end{equation}
\begin{equation}
  \frac{dR}{dt} = \gamma I
  \label{eq:R_with_lockdown}
\end{equation}
\begin{equation}
  \beta(t) = \beta_{optimal} + (1 - s(t)/100)
  \label{eq:beta_time_varying_and_stringency_dependent}
\end{equation}

@@

\begin{equation}
  \beta_{optimal} = 0.302941346409129
  \label{eq:beta_optimal_with_lockdown}
\end{equation}
\begin{equation}
  \gamma_{optimal} = 0.07020572542840187
  \label{eq:gamma_optimal_with_lockdown}
\end{equation}
\begin{equation}
  R_{0_{optimal}} = \frac{\beta_{optimal}}{\gamma_{optimal}} = 4.315051864510377
  \label{eq:r0_with_lockdown}
\end{equation}
\begin{equation}
  cost = L_{\delta = 1}(S, \hat{S}) + L_{\delta = 1}(I, \hat{I}) + L_{\delta = 1}(R, \hat{R}) = 68015990.34546319
  \label{eq:cost_with_lockdown}
\end{equation}

See figure \ref{fig:sir_with_lockdown_IND} to see how the model compares with the real data. The model's fit is better, as the $cost = 68015990.34546319$ is $1933748.65230152$ less than the model without lockdown.

@@

\subsection*{Results}
% https://www.nature.com/articles/s41598-020-79147-8 (see this to write results)
Using the simple SIR model (\ref{eq:S_without_lockdown}, ~\ref{eq:cost_without_lockdown}) to model the disease dynamics we get:
\begin{figure}[H]
  \centering
  \caption{SIR Model with lockdown for India}
  \begin{subfigure}[t]{\textwidth}
    \centering
    \includegraphics[scale=0.5]{images/SIR_model_IND.eps}
    \caption{SIR model with lockdown}
    \label{fig:SIR_mode_IND}
  \end{subfigure}
  \begin{subfigure}[t]{\textwidth}
    \centering
    \includegraphics[scale=0.50]{images/SIR_model_infections_IND.eps}
    \caption{Infections modelled with SIR model}
    \label{fig:SIR_model_infections_IND}
  \end{subfigure}
\end{figure}

>>>>

In this paper we want to optimize the government policies regarding stringency. Therefore, alongside epidemiological data, we use the measures of globally comparable government responses~\cite{Hale2021}. We use the simple SIR model without vital dynamics~\cite{Hethcote1989, Hethcote2008, ALLEN2017128} as it is assumed that the time scale is short enough so that can be neglected~\cite{Cooper2020}. By lesioning the model, as opposed to proposing a new mathematical model with more specialized compartments to more accurately represent the actual environment~\cite{Bjrnstad2020, Mwalili2020}, we can effectively address the real-world conditions and propose a solution that is both effective and extendable. The current model accounts for the recovery reached through vaccination~\cite{Marinov2022, MaurciodeCarvalho2023, Thater, Turkyilmazoglu2022, YALADANDA2022101052}. However, the study has limitations; the deterministic SIR model fails to account for chance in disease spread and lacks confidence intervals on results and while stochastic models incorporate chance, they are typically more challenging to analyze than their deterministic counterparts~\cite{Hethcote2008}.

In order to capture competing costs within the environment and achieve a balance between health and economic outcomes, we intend to employ reinforcement learning~\cite{Nguyen2022, Bastani2021, Song2022, Ohi2020, PADMANABHAN2021102676}. Not only does the formulation of the model deal better with competing costs, but it also offers a more transparency behind the reasoning of the decisions being made in such circumstances. When we conceptualize our problem as a reinforcement learning task, an agent is tasked with making decisions in an environment with the aim of optimizing cumulative rewards. This approach relies on trial and error, in contrast to dynamic programming, which assumes complete knowledge of the environment~\cite{Lavet2018}. We make the use of deep reinforcement learning, which is an advancement to reinforcement learning as it allows to reduce the input dimensionality and normalize the input~\cite{Lavet2018, Arulkumaran2017, Henderson_Islam_Bachman_Pineau_Precup_Meger_2018, Mnih2015}. While our model is more transparent in terms of its dependency on the reward policy defined, it has its limitations. A universal optimal policy may not suit diverse socioeconomic contexts due to variations in healthcare resources and economic vulnerabilities across countries, regions, or cities and a comprehensive consideration of decision factors, extending beyond pure reinforcement learning results is needed~\cite{Song2022, Dunn2017, Demir2006}.

\subsection*{Mathematical Formulation}
In SIR-RL, we use a compartmental model to model infectious disease. We iteratively develop this model to fit the actual data better. In an SIR model people of the population are divided based on whether they are yet to come into contact with an infected person (Susceptible), are infectious themselves (Infectious), or have recovered from the infection (Recovered). These compartments create the SIR model which can be represented as follows:

@@

To estimate the parameters $\beta$ and $\gamma$ for India, based on the data from May, 2020, to October, 2022, we simply define the cost to calibrate the model using the huber loss~\cite{huberloss}:

@@

\begin{equation}
  \beta_{optimal} = 0.04208479828695971
  \label{eq:beta_optimal_without_lockdown}
\end{equation}
\begin{equation}
  \gamma_{optimal} = 0.02388356686032017
  \label{eq:gamma_optimal_without_lockdown}
\end{equation}
\begin{equation}
  R_{0_{optimal}} = \frac{\beta_{optimal}}{\gamma_{optimal}} = 1.7620817917644795
  \label{eq:r0_without_lockdown}
\end{equation}
\begin{equation}
  cost = L_{\delta = 1}(S, \hat{S}) + L_{\delta = 1}(I, \hat{I}) + L_{\delta = 1}(R, \hat{R}) = 85051490.53250012
  \label{eq:cost_without_lockdown}
\end{equation}
See figure \ref{fig:sir_without_lockdown_IND} to see how the model compares with the real data.

@@

\begin{equation}
  \frac{dS}{dt} = -\beta (1 - (s(t)/100)) \frac{S I}{N}
  \label{eq:S_with_lockdown}
\end{equation}
\begin{equation}
  \frac{dI}{dt} = -\beta (1 - (s(t)/100)) \frac{S I}{N} - \gamma I
  \label{eq:I_with_lockdown}
\end{equation}
\begin{equation}
  \frac{dR}{dt} = \gamma I
  \label{eq:R_with_lockdown}
\end{equation}

@@

\begin{equation}
  \beta_{optimal} = 0.4013340889432941
  \label{eq:beta_optimal_with_lockdown}
\end{equation}
\begin{equation}
  \gamma_{optimal} = 0.09017476605499258
  \label{eq:gamma_optimal_with_lockdown}
\end{equation}
\begin{equation}
  R_{0_{optimal}} = \frac{\beta_{optimal}}{\gamma_{optimal}} = 4.450625230328213
  \label{eq:r0_with_lockdown}
\end{equation}
\begin{equation}
  cost = L_{\delta = 1}(S, \hat{S}) + L_{\delta = 1}(I, \hat{I}) + L_{\delta = 1}(R, \hat{R}) = 98438821.45587364
  \label{eq:cost_with_lockdown}
\end{equation}

Finally, an additional flow from the susceptible population to the can be shown by adding a vaccination rate $\nu$ in the model.

\begin{equation}
  \frac{dS}{dt} = -\beta (1 - (s(t)/100)) \frac{S I}{N} - \nu S
  \label{eq:S_with_lockdown_and_nu}
\end{equation}
\begin{equation}
  \frac{dI}{dt} = \beta (1 - (s(t)/100)) \frac{S I}{N} - \gamma I
  \label{eq:I_with_lockdown_and_nu}
\end{equation}
\begin{equation}
  \frac{dR}{dt} = \gamma I + \nu S
  \label{eq:R_with_lockdown_and_nu}
\end{equation}

Optimizing these equations with the Nelder-Mead method we get:

\begin{equation}
  \beta_{optimal} = 0.40897034072952304
  \label{eq:beta_optimal_with_lockdown_and_nu}
\end{equation}
\begin{equation}
  \gamma_{optimal} = 0.09196829370123338
  \label{eq:gamma_optimal_with_lockdown_and_nu}
\end{equation}
\begin{equation}
  \nu_{optimal} = 2.9044029843851936e-05
  \label{eq:nu_optimal_with_lockdown_and_nu}
\end{equation}
\begin{equation}
  R_{0_{optimal}} = \frac{\beta_{optimal}}{\gamma_{optimal}} = 4.446862329077204
  \label{eq:r0_with_lockdown_and_nu}
\end{equation}
\begin{equation}
  cost = L_{\delta = 1}(S, \hat{S}) + L_{\delta = 1}(I, \hat{I}) + L_{\delta = 1}(R, \hat{R}) = 94636860.38436058
  \label{eq:cost_with_lockdown_and_nu}
\end{equation}

However, as observed by the value of $\nu_{optimal}$ which is almost negligible and the overestimation of infected individuals in \ref{fig:SIRV_model_with_lockdown_infections_IND} suggests that $\nu$ might be varying with time, i.e., the flow of people from susceptibility to direct recovery changes with a time dependent vaccination rate. Therefore, using the $\beta_{optimal}$ and $\gamma_{optimal}$ from \ref{eq:beta_optimal_with_lockdown_and_nu} and \ref{eq:gamma_optimal_with_lockdown_and_nu}, we optimize the value of $\nu$ given a sub-interval of $[{start}, {start} + {window}]$ where the ${window} = 20$. Here, ${window}$ is a hyperparameter that can be changed so as to reduce the cost function.

\begin{equation}
  \frac{dS}{dt} = -\beta_{optimal}  (1 - (s(t)/100))  \frac{S I}{N} - \nu S
  \label{eq:S_with_lockdown_and_nu_calc_nu}
\end{equation}
\begin{equation}
  \frac{dI}{dt} = \beta_{optimal}  (1 - (s(t)/100)) \frac{S I}{N} - \gamma_{optimal} I
  \label{eq:I_with_lockdown_and_nu_calc_nu}
\end{equation}
\begin{equation}
  \frac{dR}{dt} = \gamma_{optimal} I + \nu S
  \label{eq:R_with_lockdown_and_nu_calc_nu}
\end{equation}

This model gives us how $\nu$ varies with time.

\begin{figure}[H]
  \centering
  \caption{$\nu$ Varying with Time}
  \includegraphics[width=\linewidth]{images/nu_varying_with_time_IND.eps}
  \label{fig:nu_varying_with_time_IND}
\end{figure}

Therefore, using these values we finally, recompute $\beta_{optimal}$ and $\gamma_{optimal}$ by supplying them into the equations below:

\begin{figure}[H]
\begin{tikzpicture}[node distance=1cm, auto,
  >=Latex, 
  every node/.append style={align=center},
  int/.style={draw, minimum size=1cm}]
 \node [int] (S)             {$S$};
 \node [int, right=of S] (I) {$I$};
 \node [int, right=of I] (R) {$R$};
 \coordinate[right=of I] (out);
 \path[->] (S) edge node {$\lambda$} (I)
           (I) edge node {$\gamma$} (out)
           (S) edge[bend left=45] node[above] {$\nu(t)$} (R); % Arrow from S to R with text \nu(t)
\end{tikzpicture}
\end{figure}

\begin{equation}
  \frac{dS}{dt} = -\beta  (1 - (s(t)/100))  \frac{S I}{N} - \nu(t) S
  \label{eq:S_with_lockdown_and_time_varying_nu}
\end{equation}
\begin{equation}
  \frac{dI}{dt} = \beta  (1 - (s(t)/100)) \frac{S I}{N} - \gamma I
  \label{eq:I_with_lockdown_and_time_varying_nu}
\end{equation}
\begin{equation}
  \frac{dR}{dt} = \gamma I + \nu(t) S
  \label{eq:R_with_lockdown_and_time_varying_nu}
\end{equation}
\begin{equation}
  \beta_{optimal} = 0.5129414911377119
  \label{eq:beta_optimal_with_lockdown_and_time_varying_nu}
\end{equation}
\begin{equation}
  \gamma_{optimal} = 0.12716245167626106
  \label{eq:gamma_optimal_with_lockdown_and_time_varying_nu}
\end{equation}
\begin{equation}
  R_{0_{optimal}} = \frac{\beta_{optimal}}{\gamma_{optimal}} = 4.033749620081199
  \label{eq:r0_with_lockdown_and_time_varying_nu}
\end{equation}
\begin{equation}
  cost = L_{\delta = 1}(S, \hat{S}) + L_{\delta = 1}(I, \hat{I}) + L_{\delta = 1}(R, \hat{R}) = 29742589.698910963
  \label{eq:cost_with_lockdown_and_time_varying_nu}
\end{equation}

See (\ref{fig:SIR_model_IND}, ~\ref{fig:comparing_costs_IND}) to see how the different models compare against each other.

@@

\subsection*{Results}
% https://www.nature.com/articles/s41598-020-79147-8 (see this to write results)
Using the simple SIR model (\ref{eq:S_without_lockdown}, ~\ref{eq:cost_without_lockdown}) to model the disease dynamics we get:
\begin{figure}[H]
  \centering
  \caption{SIR Model with lockdown for India}
  \begin{subfigure}[t]{\textwidth}
    \centering
    \includegraphics[scale=0.5]{images/SIR_model_IND.eps}
    \caption{SIR model with lockdown}
    \label{fig:SIR_model_IND}
  \end{subfigure}
  \begin{subfigure}[t]{\textwidth}
    \centering
    \includegraphics[scale=0.50]{images/SIR_model_infections_IND.eps}
    \caption{Infections modelled with SIR model}
    \label{fig:SIR_model_infections_IND}
  \end{subfigure}
\end{figure}
